---
title: "Mtcars data survey"

output: html_document
---
_Data Science / Regression Models / Course Project_   
_Andrey Komrakov_  
_May 27 2016_  
_Source files [https://github.com/kolfild26/regmodel.git](https://github.com/kolfild26/regmodel.git)_

### Abstract

This survey is aimed to explore the Motor Trend Car Road Tests data (_Mtcars_). We are investigating the influence of a transmission type (automatic or manual) on _MPG_ - the miles per gallon, taking into account the other car characteristics (_Number of cylinders, Gross horsepower,Rear axle ratio, Weight, Number of forward gears, etc._). 

We are trying to answer the following questions: 

1. Is an automatic or manual transmission better for MPG
2. Quantify the MPG difference between automatic and manual transmissions

We will use the _**R**_ language for the data processing, getting statistics and the linear model creation.

### Exploratory analysis

```{r settings, echo=FALSE,results='hide', message=F, warning=F}
library(ggplot2)
library(GGally)
library(car)
library(knitr)
data(mtcars)
Sys.setlocale("LC_TIME", "English")
```
First, look at the dataset we are going to work with.
```{r}
head(mtcars,5)
```

```{r, results='hide'}
range(mtcars$mpg); mean(mtcars$mpg)
```

```{r, echo=FALSE}
rng <- range(mtcars$mpg)
m <- round(mean(mtcars$mpg),1)
```

We see that the variable of intererst (outcome) **mpg** is of a number type. It varies between [`r rng`] and have a mean of `r m`.
And **am** is a factor binary variable - _0_ - automatic, _1_ - manual.

Now, let's making _t-test_ diagnostics to compare two trends (manual / automatic transmission) and find out, if there is any significant difference between them. 

```{r }
t.test(mtcars[mtcars$am==1,]$mpg , mtcars[mtcars$am == 0,]$mpg, alternative="two.sided")[3]
```
```{r , echo=FALSE}
tt <- t.test(mtcars[mtcars$am==1,]$mpg , mtcars[mtcars$am==0,]$mpg, alternative="two.sided")[3]
pv <- round(as.numeric(tt),3)
```

A small _p-value_ (typically $\leq 0.05$) indicates strong evidence against the null hypothesis. Since we get _p-value_ = `r pv`, we can assume the significant influence of a transmissiom type on the miles per gallon characteristic. Also, the same can be seen from the plot (see Appendix _picture 1._).

Let's go further and check this hypothesis based on the fact that we have more than one variable which can change an outcome.

### Multivariable modeling

First, find the variables which have a significant (greather than) correlation with _**mpg**_:

```{r }
corcoeff <- cor(mtcars$mpg, mtcars)
corcoeff[ ,abs(corcoeff) > 0.5][-1]
```
```{r, echo=FALSE}
vrbls <- rownames(as.data.frame(corcoeff[ ,abs(corcoeff) > 0.5][-1]))
```

So, _**`r vrbls`**_ might be a basis for a linear model.  
Check that the other variables do not tell us more about _**mpg**_ variance. We do this through the _**anova()**_ function which can compare the different linear models. based on their impacts in the variance explanation.

```{r}
fit01 <- lm(mpg ~ cyl + disp + hp + drat + wt + vs + factor(am),data = mtcars)
fit02 <- lm(mpg ~ cyl + disp + hp + drat + wt + vs + factor(am) + qsec + gear + carb,data = mtcars)
anova(fit01, fit02)
```

According to the _p-value_ interpretation ($> 0.05$) we can reject $H_0$, and conclude that _**qsec,gear,carb**_ are not significant in terms of variance explanation

Remember, _**anova()**_ implies the normatily of the residuals.

```{r}
c(shapiro.test(fit01$residuals)$p,shapiro.test(fit02$residuals)$p)
```
The _**p**_ values confirm normality of both model residuals, hence the anova results are comprehended.

Now, when we found the scope of parameters which vary _**mpg**_ the most, we can test the different factor conbinations to find out whether it's possible to shrink the scope of variables in the model.

Again, through _**anova()**_ we see that the model _**lm(mpg ~ factor(am) + wt + cyl,data = mtcars)**_ explains the most part of the _**mpg**_ variance.
```{r}
fit0 <- lm(mpg ~ factor(am) ,data = mtcars)
fit1 <- lm(mpg ~ factor(am) + wt ,data = mtcars)
fit2 <- lm(mpg ~ factor(am) + wt + cyl,data = mtcars)
fit3 <- lm(mpg ~ factor(am) + wt + cyl + hp,data = mtcars)
fit4 <- lm(mpg ~ factor(am) + wt + cyl + hp + disp,data = mtcars)
fit5 <- lm(mpg ~ factor(am) + wt + cyl + hp + disp + drat,data = mtcars)
fit6 <- lm(mpg ~ factor(am) + wt + cyl + hp + disp + drat + vs,data = mtcars)
```

```{r}
residM <- cbind(resid(fit0),resid(fit1),resid(fit2),resid(fit3),
                resid(fit4),resid(fit5),resid(fit6))
apply(residM,2, function(x) shapiro.test(x)$p.value )
```
All the residuals are normally distributed.

```{r}
anova(fit0,fit1,fit2,fit3,fit4,fit5,fit6)[6]
```

A residuals vs. fits plot (see Appendix _picture 2_) visually confirm that the _**lm()**_ function is being applied properly (no visible unexplained variance).  

```{r, echo=FALSE }
r_sq <- round(summary(fit2)$r.squared,2)
```
```{r, results='hide'}
summary(fit2)$r.squared
```
According to $R^2$ criteria, our model explains `r r_sq`% of a total variance.

From the model summary it can be easily seen that the difference between the manual and automatic transmision in their influence in _**mpg**_ is significant in framework of our model.

```{r}
summary(fit2 <- lm(mpg ~ factor(am) + wt + cyl - 1 ,data = mtcars))$coeff
```

```{r, echo=FALSE}
tr_dif <- round(summary(fit2)$coeff[2] - summary(fit2)$coeff[1],2)
```

### Conclusion

Having proceed the comparison between two types of transmission we detected a possible difference in thier influence on _**mpg**_. Futher, the linear model approved that the same takes place even in presence of the other factors.  
The _**lm()**_ analysis also gave as an estimation of that difference. _**Mpg**_ for the cars with manual transmission is greater on **`r tr_dif`** than for the cars with automatic transmission. Thus manual transmission is better than automatic.

### Appendix

picture 1.

```{r }
ggplot(mtcars, aes(am, mpg, fill = factor(am)) ) +
     geom_boxplot(data = mtcars, aes(x=factor(am, labels = c("Automatic", "Manual")), mpg)) +
     geom_point  (data = mtcars, aes(x=factor(am, labels = c("Automatic", "Manual")), mpg))
```

picture 2.

```{r}
par(mfrow = c(2,2))
plot(lm(mpg ~ am + wt + cyl,data = mtcars))
```